The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2023-09-19 paper.tex
last major changes from Konrad

% For one author:
%\author{\inits{N.}\fnms{Name1} \snm{Surname1}\ead[label=e1]{first@somewhere.com}}
%\address{Department first, \orgname{University or Company name},
%Abbreviate US states, \cny{Country}\printead[presep={\\}]{e1}}
% Höffner, Heuschkel, Schmiedel, Penne, Fritzsch, Ludwig, Mohaupt, Labudde, Uciteli
% TODO: overleaf
% Two or more authors:
\begin{aug}
\author[B,C]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e2]{marie.heuschkel@hs-mittweida.de}}
\author[A,C]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
\thanks{Corresponding author. \printead{e1}.}
}
\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e10]{dirk.labudde@hs-mittweida.de }}
\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e11]{alexander.uciteli@imise.uni-leipzig.de}}
\address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
Saxony, \cny{Germany}\printead[presep={\\}]{e1,e11}}
\address[B]{\orgname{Hochschule Mittweida},
Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e10}}
\address[C]{equal contribution}
\end{aug}

@@

\begin{abstract}
The Anthropological Notation Ontology (ANNO) allows the systematic and standardized classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for the derivation of different phenotypes of humans in forensic and historical anthropology.
ANNO consists of two components:
ANNOdc, a domain-core ontology providing core entities such as basic anatomical categories, and ANNOds, a domain-specific ontology used for annotating structures of the human skeleton.
ANNO is integrated into AnthroWorks3D, a photogrammetry pipeline and application for the creation and analysis of 3D-models of human skeletal remains.
The integration is based on the three-ontology method with the General Formal Ontology as the top level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are processed dynamically.
ANNO supports the analysis of skeletal and bone finds in forensic and historical anthropology, facilitating the standardization of data annotation and ensuring accurate preservation of information for posterity.
\end{abstract}

@@

%%%%%%%%%%% The article body starts:
% TODO notes from meeting, einarbeiten:
% 3 Teile Kernmodule mit Use Cases, weitere Komponenten durch Community

@@

\begin{table}[b]
  \centering
  \caption{Classes of ANNOdc along with exemplary ANNOds subclasses as well as interlinks to the FMA and the GFO top-level ontology.
  \textsuperscript{i}\anno{LO} is the line between \anno{Lambda} and \anno{Opisthion}, the \emph{occipital saggittal arc}.}
  \label{tab:core}
  \begin{tabulary}{\textwidth}{llLl}
    \toprule
	ANNOdc class					&ANNOds Examples				&FMA equivalent / ID						&GFO superclass	\\
	\midrule
	\anno{AnatomicalEntity}			&						&Physical anatomical entity \fma{61775}\\
	\anno{AnatomicalStructure}		&\anno{Skeleton}		&Material anatomical entity	\fma{67165}		&\gfo{MaterialObject}\\							
	\anno{SpatialAnatomicalEntity}	&						&immaterial anatomical entity \fma{67112}	&\gfo{SpaceEntity}\\
	\anno{AnatomicalSpace}			&						&											&\gfo{SpaceRegion}\\
	\anno{AnatomicalSurface}		&						&Anatomical surface \fma{24137}			&\gfo{SurfaceRegion}\\
	\anno{AnatomicalLine}			&\anno{LO}\textsuperscript{i}	&Anatomical line \fma{9657}	&\gfo{LineRegion}\\
	\anno{AnatomicalPoint}			&						&Anatomical point \fma{9658}							&\gfo{PointRegion}\\
	\anno{MeasurementPoint}			&\anno{Lambda}\\%, \anno{Ophistion}\\
	\anno{OrientationPoint}			&\anno{Orbitale}\\
	\anno{AnatomicalPlane}			&\anno{FrontalPlane}	&Anatomical plane \fma{242982}\\
	\anno{AnatomicalAxis}		 	&\anno{SagittalAxis}\\
	%\anno{Skeleton}					&						&Osseous skeleton \fma{23875}\\
	\anno{BoneStructure}\\
	\anno{Bone}						&\anno{Mandibula}				&Bone organ \fma{5018}\\
	\anno{BonePart}					&\anno{ArcusAlveolaris}	&Related to Segment of bone organ \fma{281808}, Zone of bone organ \fma{10483}\\
	\anno{BoneCompound}				&\anno{Cranium}				&Comparable to union of Skeletal system \fma{23881} and subdivision of skeletal system \fma{85544}\\
	\anno{ToothStructure}\\%			&\anno{DentesPermanentes}\\%		&Permanent teeth (\fma{75152}\\%; TA2ID:913)\\
	\anno{Tooth}					&\anno{DensCaninus}			&Tooth \fma{12516}\\
	\anno{ToothPart}\\%\fma{56481}; TA2ID:925\\
	\anno{Phenotype}				&\anno{SexGiles19671}			&											&\gfo{Attributive}\\
	\anno{AnatomicalProperty}		&						&													&\gfo{Attributive}\\
	%\anno{AnatomicalProperty}		&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape &\gfo{Attributive}\\
	\anno{RelativeAnatomicalLocation}&\anno{Sinister}	&Anatomical qualitative coordinate \fma{30346}	&\gfo{Relator}\\
	%\anno{RelativeAnatomicalLocation}&superior, lateral, medial, posterior	&Anatomical qualitative coordinate	&\gfo{Relator}\\
\bottomrule
\end{tabulary}
\end{table}

@@

\subsection{Validation}
When transforming the input sheet to ANNOds, SMOG~\citep{smog} executes some basic validation heuristics, such as warning on undefined references.
\todo{pyshacl}

@@

\todo{
Import der Ontologiedaten in Anwendung in Form von JSON
Anpassung der bisherigen Anwendungsstruktur an Attribute der Ontologie
Zuordnung von Objekten (Markierungen, Messungen) in Anwendung zu Objekten der Ontologie
Import und (teil-)automatisierte Berechnung der Diskriminanzfunktionen (bspw. Geschlechterbestimmung)
}

@@

\aw{}, see \cref{fig:aw}, is a tool that combines user-friendly techniques of photogrammetry with insights from user experience research and knowledge from game development and offers a procedure to create 3D skeletal models serving as  digital twins  which can subsequently be examined virtually.
This facilitates anthropological work to be location-independent and parallel without exercising wear and tear on the skeletal material.
The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
%
We use our ontological architecture (top level ontology, domain core ontology, domain specific ontology) to integrate ANNO into \aw{} based on the three-ontology method~\cite{threeontologymethod}.
%This method, as outlined in~\cite{threeontologymethod}, aided us in designing a tool for annotating human skeletons and identifying phenotypes. 
%We use ANNOdc as the Task Ontology and ANNOds as the Domain Ontology.
One of the advantages of the three-ontology method is that the software only needs to implement access to the entities (classes and properties) of ANNOdc ontology, whereas the classes of ANNOds are processed dynamically.
%\aw{} leverages ANNOdc as an interface to dynamically access the entities of the ANNOds, such as listing all subclasses of a specific class like \aurl{anno}{Bone}, as explained in \cref{sec:aw}.
%
\iffalse
It also draws its advantage when space is limited.
Thus, it requires only three SLR cameras, sufficient exposure and a computer.
The examination proceeds as follows: After photogrammetry is done and the 3D model is created, the objects are imported into the software and calibrated.
For calibration, there is an individual placeholder for each bone, which must first be selected.
Then the calibrated digital twin is placed in a placeholder skeleton.
Annotations can be made in the detailed view of each bone.
Point, line and area markings are available for selection.
Furthermore, it is possible to make measurements.
You can choose between distance, angle and circumference measurements.
After marking or measurement, an identification number is assigned to each one.
There is also a documentation about the time of the annotation as well as the name of the editor.
The marking or measurement can be classified according to certain categories (e.g. anatomical variation).
In addition, annotations are possible in a text field.
The annotations can be edited at any time afterwards, and the date of editing is recorded together with the author.
The annotations are subsequently displayed in different colors, so that overlaps can be kept apart.
A special feature of the software is the automation of measurements.
The program automatically sets pins for osteometric landmarks in a template view.
The person using the program can move these as desired.
The pins displayed have different color shades.
These differ depending on the relevance.
If the osteometric landmark is present in many measurement sections, its relevance increases and the pin appears darker.
After the pins have been roughly set, they can be refined using various views.
In the views, the individual osteometric landmarks are visible from different views, so that a fine adjustment can be made.
This is followed by the automatic measurement.
The individual values for each measurement section can be downloaded as a CSV file.
In addition, it is now possible to perform sex determinations using discriminant functions.
For this purpose, the required measurement sections can be selected.
The results are only visible in one display.
The automation allows a time saving, whereby more bones or skeletons can be examined in a shorter time.
Advantages over the previous approach are:

>>>>

\begin{aug}
\author[B,C]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e2]{marie.heuschkel@hs-mittweida.de}}
\author[A,C]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
\thanks{Corresponding author. \printead{e1}.}
}
\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e10]{dirk.labudde@hs-mittweida.de }}
\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e11]{alexander.uciteli@imise.uni-leipzig.de}}
\address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
Saxony, \cny{Germany}\printead[presep={\\}]{e1,e11}}
\address[B]{\orgname{Hochschule Mittweida},
Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e10}}
\address[C]{equal contribution}
\end{aug}

@@

\begin{abstract}
The Anthropological Notation Ontology (ANNO) allows the systematic and standardized classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for the derivation of different phenotypes of humans in forensic and historical anthropology.
ANNO consists of two components:
ANNOdc, a domain-core ontology providing core entities such as basic anatomical categories, and ANNOds, a domain-specific ontology used for annotating structures of the human skeleton.
ANNO is integrated into AnthroWorks3D, a photogrammetry pipeline and application for the creation and analysis of 3D-models of human skeletal remains.
The integration is based on the three-ontology method with the General Formal Ontology as the top level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are imported dynamically, .
ANNO supports the analysis of skeletal and bone finds in forensic and historical anthropology, facilitating the standardization of data annotation and ensuring accurate preservation of information for posterity.
\end{abstract}

@@

@@

\begin{table}[b]
  \centering
  \caption{Classes of ANNOdc along with exemplary ANNOds subclasses as well as interlinks to the FMA and the GFO top-level ontology.\\
  \textsuperscript{*} \anno{LO} is the line between \anno{Lambda} and \anno{Opisthion}, the \emph{occipital saggittal arc}.}
  \label{tab:core}
  \begin{tabulary}{\textwidth}{llLl}
    \toprule
	ANNOdc class					&ANNOds Examples				&FMA equivalent / ID						&GFO superclass	\\
	\midrule
	\anno{AnatomicalEntity}			&						&Physical anatomical entity \fma{61775}\\
	\anno{AnatomicalStructure}		&\anno{Skeleton}		&Material anatomical entity	\fma{67165}		&\gfo{MaterialObject}\\							
	\anno{SpatialAnatomicalEntity}	&						&immaterial anatomical entity \fma{67112}	&\gfo{SpaceEntity}\\
	\anno{AnatomicalSpace}			&						&											&\gfo{SpaceRegion}\\
	\anno{AnatomicalSurface}		&						&Anatomical surface \fma{24137}			&\gfo{SurfaceRegion}\\
	\anno{AnatomicalLine}			&\anno{LO}\textsuperscript{*}	&Anatomical line \fma{9657}	&\gfo{LineRegion}\\
	\anno{AnatomicalPoint}			&						&Anatomical point \fma{9658}							&\gfo{PointRegion}\\
	\anno{MeasurementPoint}			&\anno{Lambda}\\%, \anno{Ophistion}\\
	\anno{OrientationPoint}			&\anno{Orbitale}\\
	\anno{AnatomicalPlane}			&\anno{FrontalPlane}	&Anatomical plane \fma{242982}\\
	\anno{AnatomicalAxis}		 	&\anno{SagittalAxis}\\
	%\anno{Skeleton}					&						&Osseous skeleton \fma{23875}\\
	\anno{BoneStructure}\\
	\anno{Bone}						&\anno{Mandibula}				&Bone organ \fma{5018}\\
	\anno{BonePart}					&\anno{ArcusAlveolaris}	&Related to Segment of bone organ \fma{281808}, Zone of bone organ \fma{10483}\\
	\anno{BoneCompound}				&\anno{Cranium}				&Comparable to union of Skeletal system \fma{23881} and subdivision of skeletal system \fma{85544}\\
	\anno{ToothStructure}\\%			&\anno{DentesPermanentes}\\%		&Permanent teeth (\fma{75152}\\%; TA2ID:913)\\
	\anno{Tooth}					&\anno{DensCaninus}			&Tooth \fma{12516}\\
	\anno{ToothPart}\\%\fma{56481}; TA2ID:925\\
	\anno{Phenotype}				&\anno{SexGiles19671}			&											&\gfo{Attributive}\\
	\anno{AnatomicalProperty}		&						&													&\gfo{Attributive}\\
	%\anno{AnatomicalProperty}		&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape &\gfo{Attributive}\\
	\anno{RelativeAnatomicalLocation}&\anno{Sinister}	&Anatomical qualitative coordinate \fma{30346}	&\gfo{Relator}\\
	%\anno{RelativeAnatomicalLocation}&superior, lateral, medial, posterior	&Anatomical qualitative coordinate	&\gfo{Relator}\\
\bottomrule
\end{tabulary}
\end{table}

@@

%\subsection{Validation}
%When transforming the input sheet to ANNOds, SMOG~\citep{smog} executes some basic validation heuristics, such as warning on undefined references.
% pyshacl mit metaklassen wäre möglich

\subsection{Resolving incongruities}

\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{img/corpus_ossis_pubis.png}
\caption{The \emph{Corpus ossis pubis} (pubic body) forms the proximal part of the \emph{Os pubis} and includes the anterior part of the \emph{Acetabulum} as well as the \emph{Eminentia iliopubica}, which is a small elevation on the \emph{Corpus ossis pubis}.
According to the \emph{TA} and others~\citep{ta2,schemamann,anatomylexicon2007,drake2019gray,anatomylexicon,prometheus2,anatomie}, this landmark is located in the region of the \emph{Facies symphysialis}, which is defined here as the \emph{Pars symphysialis}.
}\label{fig:corpusossispubis}
\end{figure}

The domain experts reconciled conflicting literature to arrive at a consistent and logical result, for example for the \emph{Corpus ossis pubis}, see \cref{fig:corpusossispubis}:
Since the Corpi of the \emph{Os ischii} and \emph{Os ilii} are located in the area of the \emph{Acetabulum},
it is logical to place the \emph{Corpus ossis pubis} in this region as well~\citep{graysanatomy,waldeyer,allgemeineanatomie,datacollection,romanianmandible}, thereby forming the \emph{Acetabulum} from all Corpi (\emph{Corpus ossis ilii}, \emph{Corpus ossis ischii}, and \emph{Corpus ossis pubis}).
The \emph{Pars symphysialis ossis pubis} has ventral-medially the \emph{Facies symphysialis}, which connects the paired \emph{Ossa pubes} of the pelvic bones.
In the \emph{Pars symphysialis}, the \emph{Ramus inferior ossis pubis} and \emph{Ramus superior ossis pubis} merge into each other.
This redefinition closes the resulting gap in the definition of the area between the Ramus \emph{inferior / superior} ossis pubis.
%
ANNO also corrects the commonly missing differentiation between \emph{Cranium} and \emph{Mandibula} in English literature, with no Latin equivalent for the overall term \enquote{skull}.
Instead, \anno{Cranium} is established as a synonym for the skull~\citep{allgemeineanatomieen,prometheus2}.
\paragraph{}
Further contributions of ANNOds include:
\begin{itemize}
\item Consistent Latin declension (including in plural).
\item Transparency through detailed source references; for measurement distances, origin works and reference works are indicated by dividing into primary and core sources.
\item Dissected group structures.
\item Redefinitions such as joint surfaces, which are often not named in anatomy literature but are anthropologically relevant, and osteometric measurement points outside the skull, which are not established in the literature.
\end{itemize}

@@

@@

\aw{}, see \cref{fig:aw}, is a German tool that combines user-friendly techniques of photogrammetry with insights from user experience research and knowledge from game development and offers a procedure to create 3D skeletal models serving as digital twins which can subsequently be examined virtually.
This facilitates anthropological work to be location-independent and parallel without exercising wear and tear on the skeletal material.
The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
%
We use our ontological architecture (top level ontology, domain core ontology, domain specific ontology) to integrate ANNO into \aw{} based on the three-ontology method~\cite{threeontologymethod}.

\begin{figure}[h]
\includegraphics[width=0.7\textwidth]{img/json.png}
\caption{Left: ANNO converted to the \aw{} JSON input format. Right: Bone selection in the imported hierarchy.}
\label{fig:json}
\end{figure}

%This method, as outlined in~\cite{threeontologymethod}, aided us in designing a tool for annotating human skeletons and identifying phenotypes.
%We use ANNOdc as the Task Ontology and ANNOds as the Domain Ontology.
One of the advantages of the three-ontology method is that the software only needs to implement access to the entities (classes and properties) of ANNOdc ontology, whereas the classes of ANNOds are processed dynamically, as shown in \cref{fig:json}.
%\aw{} leverages ANNOdc as an interface to dynamically access the entities of the ANNOds, such as listing all subclasses of a specific class like \aurl{anno}{Bone}, as explained in \cref{sec:aw}.
%
In addition, it is now possible to perform sex determinations using discriminant functions, as shown in \cref{fig:function}.

\begin{figure}[h]
\includegraphics[width=0.4\textwidth]{img/function.png}
\caption{Sex determination in \aw{} using the imported ANNO discriminant functions.}
\label{fig:function}
\end{figure}

\iffalse
It also draws its advantage when space is limited.
Thus, it requires only three SLR cameras, sufficient exposure and a computer.
The examination proceeds as follows: After photogrammetry is done and the 3D model is created, the objects are imported into the software and calibrated.
For calibration, there is an individual placeholder for each bone, which must first be selected.
Then the calibrated digital twin is placed in a placeholder skeleton.
Annotations can be made in the detailed view of each bone.
Point, line and area markings are available for selection.
Furthermore, it is possible to make measurements.
You can choose between distance, angle and circumference measurements.
After marking or measurement, an identification number is assigned to each one.
There is also a documentation about the time of the annotation as well as the name of the editor.
The marking or measurement can be classified according to certain categories (e.g. anatomical variation).
In addition, annotations are possible in a text field.
The annotations can be edited at any time afterwards, and the date of editing is recorded together with the author.
The annotations are subsequently displayed in different colors, so that overlaps can be kept apart.
A special feature of the software is the automation of measurements.
The program automatically sets pins for osteometric landmarks in a template view.
The person using the program can move these as desired.
The pins displayed have different color shades.
These differ depending on the relevance.
If the osteometric landmark is present in many measurement sections, its relevance increases and the pin appears darker.
After the pins have been roughly set, they can be refined using various views.
In the views, the individual osteometric landmarks are visible from different views, so that a fine adjustment can be made.
This is followed by the automatic measurement.
The individual values for each measurement section can be downloaded as a CSV file.
In addition, it is now possible to perform sex determinations using discriminant functions.
For this purpose, the required measurement sections can be selected.
The results are only visible in one display.
The automation allows a time saving, whereby more bones or skeletons can be examined in a shorter time.
Advantages over the previous approach are:
